---
title: "Summer-To-Do"
author: "Daniel Kaplan"
date: "May 6, 2014"
output: pdf_document
---

```{r include=FALSE}
require(ggplot2)
require(dplyr)
```

## First day

* Figure out about laptop.  Maybe Danny's office machine.
* Install git
* Install [preview version of RStudio](http://www.rstudio.com/ide/download/preview)
* Install latest version of R
* Write and make a screenflow video of how to do this. (3 minutes)
    * Review the whole process with slides, including links
    * Illustrate each step in the process
    * Maybe process this as a tutorial with embedded videos.  See [HTML5 video tag](http://www.w3schools.com/html/html5_video.asp)

## Set up a stub Jekyll site for a course

## Get direct reading of Google Spreadsheets to work

## Look at the `knitr` side-by-side display app.

See the `inst/shiny` directory on the knitr GitHub site.

Make a few example documents that people can play with.  E.g. show them a bulleted list, ask them to add an item, to create a sublist, to make a new section.  To make a new block and do some R calculation in it.

## Graphics

### simple map-drawing functions

See `SummerTasks/MapProgram.Rmd[html]`

### Look for on-line lovely graphics producing programs

See what format data needs to be in, how to upload it, and create examples of using it.  I want to open up the graphics from the pure R functions.

### Pick out some compelling interactive graphics from here

A large collection of interactive graphs: http://rcharts.io/gallery/

## Build Shiny Apps out of the mosaicManip apps

One directory for each app.

Try the shiny-Rmd for some of them.

Do this for just a few, so you'll be in a position to write new, DCF apps.

## Create a Pivot-vs-Fold app

Like the one on p.88 of *Data Transformation_ Skills of the Agile Data Wrangler Presentation.pdf* , but interactive.

## Country data

Create a table of synonymous country names and codes.  `CaseStudies/Countries`

* `countrycode_data` in `countrycode` package
* The map package (used in 2013) has a table 

## Scrape the Zip-Code data

They are in URLs like <http://www.brainyzip.com/state/zip_newyork.html>

Sort out what's useful in `CaseStudies/ZipCodes/zip_codes_states.csv` which is from <http://notebook.gaslampmedia.com/download-zip-code-latitude-longitude-city-state-county-csv/>.  Note that leading zeros have been dropped from the zip code.  A data cleaning exercise?

## Scrape the County data

<http://quickfacts.census.gov/qfd/download_data.html> has a database

<http://censtats.census.gov/usa/usa.shtml>

## Bring the Cherry Blossom race up to date

See `CaseStudies/CherryBlossomRace/notes.Rmd`

## Federal Election Commission case study

Hadley provides software for reading the file.  Do this (out of the project --- it's too large to include on github) and then extract a manageable subset.

## Crime Data

As in [Houston](http://www.houstontx.gov/police/cs/stats2.htm)

Break-down crime by day-of-week and offense.  Quick start: `ggmap::crime`

[FBI data](http://www.fbi.gov/about-us/cjis/ucr/ucr-program-data-collections#National)

This one uses: 

* `group_by()` see summarize()
* `summarize()` counting events by offense, day, hour, month, etc.
* `filter()` pulling out specific crimes
* `mutate()` turning the number in a day into the percent for that day
* `select()` getting rid of extraneous variables

```{r eval=FALSE}
f <- ggmap::crime
ff <- select(f, hour, day, offense, month)
gff <- group_by(ff, hour)
hcount <- summarize(gff, n=n())
ggplot(hcount, aes(x=hour,y=n)) + geom_line(color="blue",width=4)
crimeCount <- group_by(ff,offense,hour) %.% tally()
ggplot(crimeCount,aes(x=hour,y=n,fill=offense)) + geom_bar(stat='identity',position=position_stack(width=.9)) +facet_wrap(~day)
# Do all three types: position_fill, position_dodge, position_stack
ggplot(filter(propCrimeCount,offense%in%c('rape','murder')),
       aes(x=hour,y=p,fill=offense)) + geom_bar(stat='identity',position=position_dodge(width=.9)) 
# Now reduce each to the proportion
propCrimeCount <- group_by(crimeCount,offense) %.% mutate(p=n/sum(n))
ggplot(filter(propCrimeCount,offense%in%c('rape','murder')),
       aes(x=hour,y=p,fill=offense)) + geom_bar(stat='identity',position=position_dodge(width=.9)) 
```

## European Temperature data

For fields and mutation, wide, long, ....

<http://www.datasciencecentral.com/group/resources/forum/topics/importing-100-years-of-climate-change-into-r>

## US Health Data

[Data site](http://www.tycho.pitt.edu/)  [NEJM article](http://www.nejm.org/doi/pdf/10.1056/NEJMms1215400)

## Apps for data operations

### Mutate



* Fixed set of example data -or- Pick a data set and create the mutation you want by writing in the expression.
* Show the head of the mutated data and the new column alongside.
* Show the statement that will do the job.
* Some quiz exercises.
    * Quiz results: Get the userID, then code the table of results for them to send in.  Or, can we arrange to send it automatically?

### Join

* Selection of kind of join
* Have a list of example data types, and 
* Selection of tables to join from the local disk
* Choose kind of join
* Choose vars to join on.   Choice from table A -> Choice from table B

### Set up a course web site with Jekyll

Resources, syllabi (perhaps multiple) that refer to the resources.

