---
title: "Brainstorming about Disasters"
date: "May 23, 2014"
output: html_document
---

Dominant tasks: 

* scatter plot  map as guide
* counts, sums, mutate (new variable)

choropleths

The `emdatr` package provides access to the International Disaster Database, EMDAT, from the Center for Research on the Epidemiology of Disasters.

```{r include=FALSE}
require(emdatr)
```

Extract the 2013 data:
```{r}
D2013 <- extract_emdat() # for 2013 by default
# Whole dataset
#    D <- extract_emdat(sample_only=FALSE)
# I saved this as a local file, for speed
load("WholeDataset.Rda")
```

Watch out.  `Killed` and `TotAffected` are integers, convert to numeric before using.

## Which type of disaster has had the most impact in terms of deaths and people affected.

```{r}
Dsize <- group_by(D,Type) %.% 
  summarize( killed=sum(as.numeric(Killed),na.rm=TRUE), 
             size=sum(as.numeric(TotAffected), na.rm=TRUE))
Dsize
```

## Number of disasters versus Year

plot out count against year.  group_by

order by count, to show the distribution of the years.

Same for total estimated damage in each year.  Order by total damage.


## Group by GDP per capita

Show what type of disaster varies for the different GDP/capita groups.

Total count in GDP group versus population in that group.

Does it vary by total GDP?

## Group by Latitude

More near the equator.


## Plot out location of extreme temperatures

1. Get the country location data --- MAKE PART OF DCF PACKAGE
2. Join location with disaster
3. Dot plot of `TotAffect` or something else about temperatures.  
   * Use to illustrate log or sqrt transformations.
   
Other levels aside from extreme temperatures: where do they happen

put border geom on top of dot plot.

But we'll save the data `as.numeric()`

## Is there a preferred month for disasters?

Break down by group/type of disaster.

## How long do disasters last?

## Cleaning or strsplit()

Turn the Start & Stop into a month

```{r eval=FALSE}
# This doesn't work: Use the date functions.
broken_up <- unlist(strsplit(D$Start,'/'))
first <- broken_up[seq(1,length(broken_up),by=3)]
second <- broken_up[seq(2,length(broken_up),by=3)]
third <- broken_up[seq(3,length(broken_up),by=3)]
```

# Get this to work

```
head(as.Date(D$Start, "%d/%b/%y")) # Doesn't work
```

SOFTWARE TO WRITE:
Function that takes a string, figures out about the format (heuristically), and depending on the flag, turns it into a POSIX date, or a month, day, or year.


## Population related

Look at number affected 

* as function of population
* as function of GDP per population

NEED A POPULATION table to merge with.  GapMinder: narrow to wide, interpolate, each year.
