---
title: "Assignment for Week Two"
author: "Data and Computing Fundamentals"
date: "Macalester College"
output:
  html_document:
    css: ~/KaplanFiles/DCF-2014/ReOrganizedJune10/CSS/DCF-style.css
    fig_caption: yes
    toc: yes
---

```{r child="/Users/kaplan/KaplanFiles/DCF-2014/ReOrganizedJune10/CSS/DCF-common.Rmd"}
```

* Groupwise means and sd.
* Counts using group_by() and n().  (Group by the levels of the variable to be counted, perhaps including another variable for subdivisions.)
* Converting counts to proportions. (Divide by sum().)
    ```{r}
    KidsFeet %>% group_by( biggerfoot,sex ) %>% 
    summarize(count=n() )
    ```
* Graphics with mPlot(), mBar(), mMap()

* Add a layer to a map, perhaps zip-code population with state boundaries.

## Deconstructing Graphics

Here's a link to a graphic.  You are to include it in your assignment write-up and then provide the answers.  (Remember, you can right-click on the link to get the URL, then paste that into your Rmd document.)

DECONSTRUCTING EXAMPLES
