Zip Code Data
===============

```{r include=FALSE}
require(XML)
require(RCurl)
``` 

To analyze the Consumer Financial Products Bureau database, it's helpful to know the population of each zip code.  Perhaps the Census Bureau provides such information, but I haven't found it yet.  In searching around on the Internet, I found a site, <http://www.brainyzip.com/state/zip_minnesota.html> that give population-by-zip code information for Minnesota.  

The web site also has information for other states.  So I thought I would try to "scrape" the site for the information I sought.

First, to get a list of the state and their populations.  That will help to double-check the zip-code data.
```{r}
stateURL <- "http://www.indexmundi.com/facts/united-states/quick-facts/all-states/population#table"
# it happens to be in the 3rd table in the document
stateData <- readHTMLTable(getURLContent(stateURL),which=3)
head(stateData)
```

Alas, the population isn't being stored as a number, even if it looks like it.  The commas cause R to treat `Value` as a categorical variable.  We can fix that:
```{r}
stateData <- transform(stateData,
                       Value=as.numeric(gsub(",","",as.character(Value))))
head(stateData)
```

Now to scrape the zip-code data from the website....


Here's how, using the `XML` and `RCurl` packages:
```{r eval=FALSE}
stateNames <- stateData$State
res <- NULL
for (state in stateNames) {
  state <- tolower(gsub(" ","",state)) # kill spaces and upper case
  theURL <- paste("http://www.brainyzip.com/state/zip_",
                  state,".html",sep="")
  tables <- readHTMLTable(getURLContent(theURL))
  tab <- tables[[3]][-1,] # drop the first entry
  names(tab) <- c("Zip","City","State","County","Population")
  res <- rbind(res,tab)
}
res <- transform(res, Population=as.numeric(as.character(Population)),
                 Zip=gsub(' zip code','', Zip))
zipCodeLocation <- droplevels(res)
# Too much space
# write.csv(zipCodeLocation,'zipCodeLocation.csv',row.names=FALSE)
save(zipCodeLocation,'zipPop.Rda')
```

Links to get more data for each zip code: <http://www.brainyzip.com/economic/71/economic_71602.html>