

## Review

* CVC/Summer2014/Activities/NCI60.Rmd.  There's only one join operation in the thing, but that's why the topic is in this week rather than week 3.

## Country Names 

Standardizing country codes: See the CVC/Summer2014 exercise: WorldPop.Rmd

Disasters --- join with population data to produce a risk factor for each state.

### Entropy of Baby Names

Turn each name count into a fraction for that year.  Then calculate the entropy over time for boys and girls separately.

```{r}
Entropy <- BabyNames %>%
  group_by( sex, year ) %>%
  mutate( frac=count/sum(count) ) %>%
  ungroup() %>%
  group_by( sex, year ) %>% 
  summarise( entropy=sum( -frac*log2(frac)))
```

Plot it out:
```{r}
ggplot( Entropy, aes(x=year, y=entropy, group=sex, color=sex ) ) +
  geom_line( )
```

Higher entropy corresponds to a greater diversity of names.  An entropy of 7 is equivalent to $2^7 = 128$ different names, each of equal likelihood.  An entropy of 10 is equivalent to $2^{10} = 1024$ different names.

### Arrange within group_by()

Or maybe it should be one of the rank functions in filter.

Within each precinct, which ward had the largest number of votes?  (Hint: Make a data table with the number of votes in each `Precinct` `Ward` combination.  Arrange this by `Precinct` and the number of votes.)

### Immigration data

In data scraping, convert the wide format to narrow.  Move to week 7.

### NCI60

Convert the wide format to narrow.

### Pull out the 50 most popular gender-neutral baby names

Then plot out a line graph to see how they change in balance over time.

