---
title: "Medicare Procedures Analysis in dplyr"
author: "Daniel Kaplan"
date: "April 21, 2014"
output: html_document
---

```{r include=FALSE}
require(dplyr)
require(mosaic) # for mPlot()
require(DCF) # for mBar()
```

Read in the three tables: `occasions`, `procedures`, and `providers`
```{r}
load('MedicareBilling.rda')
occasions_small <- occasions[1:10000,]
```

## Simple summaries

### How many providers of each qualification

The `dplyr` function `tally()` is a wrapper for `summarize()`
```{r}
quals <- arrange( 
  tally(group_by(providers,credential)),
  desc(n))
```

```{r echo=FALSE}
ggplot(data=quals,aes(x=reorder(credential,n),y=n ))+geom_bar(stat='identity',position=position_stack(width=.9)) + theme(axis.text.x=element_text(angle=60,hjust=1)) 
```

### Break down credentials according to the entity code

```{r}
quals2 <- arrange( 
  tally(group_by(providers,credential,entity_code)),
  desc(n))
```

Evidently, nurses aids are the O-type entities
```{r echo=FALSE}
ggplot(data=quals2,aes(x=reorder(credential,n),y=n ,fill=entity_code))+geom_bar(stat='identity',position=position_stack(width=.9)) + theme(axis.text.x=element_text(angle=60,hjust=1)) 
```

### Break down credentials by sex

```{r}
quals3 <- arrange( 
  tally(group_by(providers,credential,gender)),
  desc(n))
```

Evidently, nurses aids are the O-type entities
```{r echo=FALSE}
ggplot(data=quals3,aes(x=reorder(credential,n),y=n,order=reorder(gender,n),fill=gender))+geom_bar(stat='identity',position=position_stack(width=.9)) + theme(axis.text.x=element_text(angle=60,hjust=1))
```

Improve graph by plotting the proportion, and ordering it by a variable that gives the proportion.

What's the strategy here?  For each credential, find the proportion of women.  Then join that with the counts of credentials broken down by sex.  **JOIN** **GROUP**  **COUNT**

Look at the proportion in each credential.

```{r}
group_by(providers, credential) %.% 
  summarize( propF=mean(gender=='F')) -> credFprop
head(credFprop)
group_by(providers, credential, gender) %.% 
  tally() %.% inner_join(credFprop)  -> result
```

```{r echo=FALSE}
ggplot(data=result,aes(x=reorder(credential,propF),y=n ,fill=gender))+geom_bar(stat='identity',position=position_fill(width=.9)) + theme(axis.text.x=element_text(angle=60,hjust=1)) 
```





