
```{r child="/Users/kaplan/KaplanFiles/DCF-2014/CSS/DCF-common.Rmd"}
```

The `ZipGeography` data table in the DCF package gives information about each postal code ("ZIP code") in the US.  Here's an excerpt of the `r nrow(ZipGeography)` cases 

```{r echo=FALSE}
set.seed(101)
ZipGeography %>% 
  filter( State %in% c("Colorado","Minnesota","California","Maryland"),Population>0, LandArea > 0) %>% 
  sample_n( size=10 ) %>%
  select( ZIP, State, Population, LandArea ) %>%
  arrange( State, ZIP )
```

Here's a graphic showing the mean population of all the ZIP codes in each state.  

```{r echo=FALSE}
ZipGeography %>%
  select( ZIP, State, Population, LandArea) %>%
  group_by( State ) %>% 
  summarise( 
    aveZipPopulation=mean(Population, na.rm=TRUE),
    area=mean(LandArea, na.rm=TRUE)
    ) %>%
  mutate( State=reorder(State, aveZipPopulation) ) %>% 
  ggplot( aes(x=State,y=aveZipPopulation) ) +
  geom_point( aes(size=area)) +
  theme(axis.text.x=element_text(angle=60,hjust=1))
```

1. Are the `ZipGeography` data in glyph-ready form for this graphic?  
    * If so, explain which variable in `ZipGeography` is being mapped to which graphical attribute.  
    * If not, explain in words how the `ZipGeography` data might have been transfigured to create a glyph ready data table for this graphic.
2. The variable mapped to the horizontal axis is, obviously, `State`.  Explain what other variable is being used to set the scale for the horizontal axis, that is, how each level of `State` is mapped to a position on the axis.
3. What are the three guides in the graphic?
