---
title: "BRCA"
author: "Daniel Kaplan"
date: "May 29, 2014"
output: html_document
---

```{r include=FALSE}
require(dplyr)
require(ggplot2)
require(reshape2)
require(ggthemes)
```

## Resources

* [Volcano plot](http://www.r-bloggers.com/using-volcano-plots-in-r-to-visualize-microarray-and-rna-seq-results/?utm_source=feedburner&utm_medium=email&utm_campaign=Feed%3A+RBloggers+%28R+bloggers%29)

## BRCA

But keep in mind that it's a mutation in BRCA that's associated with an increased risk of cancer.  These data are **not** comparing normal and mutation.

Looking for the relationship between the BRCA gene and cancer.

Paul Overvoorde provided a file from the NCI site.  [NEED MORE DETAILS].  It gives gene expression levels in 60 subjects.  

```{r}
fileName <- "/Users/kaplan/Downloads/filesandlinks/nci60_RNA__Agilent_mRNA_log2.txt/nci60_RNA__Agilent_mRNA_log2.txt"
geneExpress <- read.table( fileName, head=TRUE )
```

```{r}
geneNames <- as.character(geneExpress$Probe)
```

Find the ones with a name indicating BRC
```{r}
keepers <- filter(geneExpress, grepl("BRC", as.character(Probe)))
nrow(keepers)
```
There are `r nrow(keepers)` such genes.

Transform from a wide format (where each column is a patient) to a narrow format (where columns are cell-line, probe, measurement).
```{r}
express <- melt(keepers,variable='Cell.Line.Name',value.name='expression')
allgenes <- melt(geneExpress,variable='Cell.Line.Name',value.name='expression' )
```

The information about the cell lines:

* Note, the punctuation is different in cellLineData than in express, so merging  doesn't work.  Fix that by changing the punctuation.
```{r}
cellLineFile <- "/Users/kaplan/KaplanFiles/DCF-2014/CaseStudies/NCI-gene/nci60_cellline_metadata.csv"
cellLineData <- read.csv( cellLineFile, head=TRUE)
# There are different punctuations in the Cell.Line.Names for express and cellLineData.
cellLineData <- mutate(cellLineData, Cell.Line.Name=sub(":",'.',Cell.Line.Name))
both <- inner_join(express , cellLineData)
bothall <- inner_join(allgenes, cellLineData)
```

Construct the table for BRCA1

```{r}
origs <- c('BRCA1','DTX3L',"BID","CALU","BFAR", "NONO")
top100 <- c('ACTG2' , 'SULF1' , 'COL3A1' , 'IGFBP5' , 'SLC24A3' , 'COL1A2' , 'LUM' , 'SPARC' , 'POSTN' , 'CXCL1' , 'A_23_P37685' , 'MGC16121' , 'UCHL1' , 'A_23_P203120' , 'ALDH2' , 'DCN' , 'DPYSL3' , 'CPXM2' , 'WISP1' , 'CDKN2A' , 'FEZ1' , 'CBS' , 'A_24_P372189' , 'CDCP1' , 'C10orf58' , 'CNN1' , 'A_32_P74409' , 'SYNPO2' , 'CDCP1' , 'WNT5A' , 'A_23_P305292' , 'KIAA0746' , 'COL1A1' , 'CDH11' , 'A_24_P299474' , 'GPR110' , 'A_23_P36226' , 'A_32_P214011' , 'A_23_P217901' , 'SPOCK1' , 'ACTA2' , 'KRT19' , 'A_24_P563966' , 'KRT17' , 'CXCL12' , 'KCNG1' , 'HSPB2' , 'VCX2' , 'FAP' , 'MRVI1' , 'TSHZ3' , 'ADAMTS4' , 'COL3A1' , 'OLFML3' , 'IFITM1' , 'RIMS2' , 'C19orf33' , 'BST2' , 'DPYSL3' , 'C11orf41' , 'VCX' , 'C5orf23' , 'CSTA' , 'ROBO1' , 'FOXD1' , 'IGFBP5' , 'CD248' , 'MX1' , 'ZNF469' , 'PLAC8' , 'ADFP' , 'APCDD1L' , 'CDH13' , 'CTHRC1' , 'SAA1' , 'A_32_P55241' , 'TMEM98' , 'COX7A1' , 'CST6' , 'A_23_P206741' , 'FAM101B' , 'A_32_P96752' , 'FKBP10' , 'BDKRB1' , 'A_24_P691826' , 'A_24_P887857' , 'VCX3A' , 'PCSK9' , 'ADAMTS5' , 'CTAG1A' , 'SIPA1L2' , 'A_24_P828496' , 'ROBO3' , 'G0S2' , 'KCNMA1' , 'ITGA8' , 'PLAC8' , 'PRAME' , 'PTGIS' , 'MAGEA2B' )
probeSet <- c(origs, top100)
res <- list()
for (probe in probeSet) {
  filter(bothall, Probe==probe) %>% 
  group_by(tissue.of.origin) %>% 
  summarize(ave=mean(expression),stderr=sd(expression)/sqrt(n())) -> tmp
  tmp$tissue.of.origin <- with(tmp,reorder(tissue.of.origin,-ave))
  res[[probe]] <- tmp
}
```

```{r}
Fres <- data.frame( probe=character(length(probeSet)), F=NA, p=NA,
                    stringsAsFactors=FALSE)
for (probe in probeSet) {
  P1 <- ggplot(data=res[[probe]], aes(x=tissue.of.origin,y=ave,color=tissue.of.origin)) 
  P2 <- P1 + 
    geom_bar(stat='identity',position='dodge',alpha=0.2,color='gray') +
    geom_errorbar(aes(ymax=ave+2*stderr,ymin=ave-2*stderr),
                  width=0.2,position=position_dodge(width=0.8))
  raw <- filter(bothall, Probe==probe)
  P3 <- P2 + geom_point(data=raw,
                        position=position_jitter(width=0.15),
                        aes(y=expression),alpha=.5)
  print(P3 + ggtitle(probe) + xlab('Tissue type') +
          ylab('Expression level (log2)') +
          theme_tufte() + coord_flip())
  tmp <- anova(lm(expression ~ tissue.of.origin, data=raw))
  print(tmp)
  Fres[which(probe == probeSet),] <- list(probe=probe, F=tmp[1,4], p=tmp[1,5] )
  }
Fres
```