
```{r include=FALSE}
require(mosaic)
require(dplyr)
require(ggplot2)
require(tidyr)
require(ScoreR)
startProblem("DCF-Assignment2")
```

```{r child="../CSS/DCF-common.Rmd"}
```

## Week 2 Drill


### Multiple Choice

There are three (simple) steps to creating a multiple-choice problem in *ScoreR*:

1. Initialization of the question
2. Creation of the individual items, typically embedded in *markdown* statements
3. Termination of the question.

Initialization is done with the `newMC()` function.

```{r}
f = newMC(totalPts=5,name="minus1") # Ordinarily, make this silent
```

The function name, in this case `f`, is arbitrary.  You will use that name to create the individual items with statements like `I(f(TRUE))` and I(f(FALSE))`, which will be embedded in text.  For instance:

Which of these is equal to $-1$?  
`r I(f(FALSE))` The value of $a x^2 + b x - c$ for all $a$, $b$, and $c$  
`r I(f(TRUE))`  $e^{i\pi}$  
`r I(f(FALSE))` $\ln(0)$  
`r I(f(FALSE))` There's no such thing as a negative number, just a negative attitude.
`r I(f(finish=TRUE))`

After the items have been defined, terminate the question by calling the function with the argument `finish=TRUE`
```{r eval=FALSE} 
I(f(finish=TRUE))
```
You may well want to use inline knitr notation --- that's up to you.

### Select from a set

What's the value of $\pi$?
```{r}
piQuestion3 = selectSet(name="piValue3",totalPts=4,
                        "2.71727"=FALSE,
                        "3.0"=FALSE,
                        "3.14159"=TRUE,
                        "6.28318"=FALSE)
```

* Let's try that again ... `r I(piQuestion3)`


### Free Text

The `freeText()` function in `ScoreR` implements free-text entry.  It's very, very simple to add a free-text question to a problem.  Just

`r I(textItem(name="question-topic",rows=2,totalPts=10))`

The name, of course, should be unique within the problem.  The optional arguments `rows` and `cols` sets the size of the text entry box.  On many browsers, this block can be resized by the user, so a small box is useful in for adding in extended text.


### Second Part

A perennial problem when writing scientific reports is figuring out
how many significant digits to report.  It's naïve to 
copy all the digits from one's calculator or computer
output; the data generally do not justify such
precision. 

Once you have a confidence interval, however, you do not need to guess
how many significant digits are appropriate.  The standard error
provides good guidance.  Here is a rule of thumb: keep two significant
digits of the margin of error and round the point estimate to the same
precision.  

For example, suppose you have a confidence interval of $1.7862 \pm
0.3624$ with 95% confidence.  Keeping the first two significant
digits of the margin of error gives $0.36$.  We'll keep the point
estimate to the same number of digits, giving altogether $1.79 \pm
0.36$.

Another example: suppose the confidence interval is $6548.23 \pm
1321$.  Keeping the first two digits of the margin of error gives
$1300$, with a corresponding margin of error of $6500 \pm 1300$.

* Suppose the computer output is $0.03234232 \pm 0.01837232$.

```{r include=FALSE}
f=newMC(totalPts=1, name="confidence.interval")
```
Using this rule of thumb, what should you report in as the
confidence interval?      
`r I(f(FALSE))` $0.3234 \pm 0.01837$     
`r I(f(FALSE))` $0.3234 \pm 0.0183$     
`r I(f(FALSE))` $0.0323 \pm 0.0184$    
`r I(f(FALSE))` $0.0323 \pm 0.018$     
`r I(f(TRUE))` $0.032 \pm 0.018$     
`r I(f(FALSE))` $0.032 \pm 0.01$     
`r I(f(FALSE))` $0.03 \pm 0.01$   
`r I(f(finish=TRUE))`


* Now suppose the computer output were $99.63742573 \pm 1.48924367$.

```{r include=FALSE}
g=newMC(totalPts=1, name="confidence.interval.2")
```
What should you report as the confidence interval?     
`r I(g(FALSE))` $100 \pm 1$     
`r I(g(FALSE))` $99 \pm 1.5$     
`r I(g(TRUE))` $99.6 \pm 1.5$     
`r I(g(FALSE))` $99.64 \pm 1.49$     
`r I(g(FALSE))` $99.647 \pm 1.489$ 
`r I(g(finish=TRUE))`

`r I(endProblem())`
