---
title: "MammalianSpecies"
author: "Daniel Kaplan"
date: "May 25, 2014"
output: html_document
---

```{r include=FALSE}
require(dplyr)
require(ggplot2)
```

The dataset `msw3-all.csv` comes from Don E. Wilson & DeeAnn M. Reeder (editors). 2005. *Mammal Species of the World. A Taxonomic and Geographic Reference* (3rd ed), Johns Hopkins University Press <http://www.departments.bucknell.edu/biology/resources/msw3/>

The original file has much information about the source of the original classification and distribution.

### Simplified Dataset

```{r eval=TRUE}
M <- read.csv("/Users/kaplan/KaplanFiles/DCF-2014/CaseStudies/Mammals/msw3-all.csv")
keep <- M[,c(1,2,6,9,11,12,13,14,15,18,26)]
M <- keep
save(M, file='speciesM.Rda')
```

### How many of each Taxon Level?

```{r}
group_by( M, TaxonLevel) %.% tally() %.% arrange(-n)
```

### Are there any distinct species with the same order, etc.




### Cleaning

What can we do with the location data?

Fixing the Years for easier analysis.

If they have a suborder, then they have an order.  True?
