---
title: "Making Maps with ggplot2, maps, and mapproj"
author: "Daniel Kaplan"
date: "May 9, 2014"
output: html_document
---

```{r include=FALSE}
library(dplyr)
library(maps)
library(mapproj)
library(ggplot2)
```

Some notes on how to make world/country/county maps.

Goal: Produce a small set of functions that we can use to streamline map-making from data.

You need the shapefiles from `maps` and the projections from `mapproj`, as well as `ggplot2` to do the drawing.

```{r}
library(dplyr) #always!
library(maps)
library(mapproj)
library(ggplot2)
```

### 1. Choose a Region 

We can use this directly.

```{r}
shapes <- map_data('world')
```

### 2. Get data 

Get data for the units of the map, e.g. countries.  These are identified under `region` in `shapes`. 

General data operations.  Let's have a few sample files.

### 3. Reconcile the codes in your map to those in the shapes file.

Write a function that compares the shape info and data to point out any missing or mismatched ones.  I think there is already functionality in the `rworldmap` package to do this.  Use that if it will work.


### 4. Merge the reconciled data with the shapefile.

I'm thinking that it might be good to provide one function that takes the shapefile and the reconciled data, merges them, calls `ggplot()` and returns a ggplot object which has no layers.

In writing the function, use syntax like this 
```{r eval=FALSE}
myMap <- makeMap(shapefile,datafile,varname)
```

> For the next items, I'm going to make a bogus dataset in the right form.  Since the `shapes$regions` are `chars`, I need to make sure that the conversion in `myData` is correct.

```{r}
countryNames <- unique(shapes$region)
myData <- data.frame( val=runif(length(countryNames)),
                      region=countryNames,stringsAsFactors=FALSE )
shapesAndData <- inner_join(shapes,myData)
```

> Now use `shapesAndData` for the plotting.

The basic `ggplot2` syntax is

```{r}
myMap <- ggplot(shapesAndData, aes(x=long,y=lat,group=group,order=order))
```

Notice, the `fill=val` argument.  `val` stands for the name of the variable to be plotted, which is coming from `datafile`.

, myData)
```
### 5. Plot out the map 

Add the layers.  Two basic types: outlines and fills.

Outlines:
```{r}
myMap + geom_polygon(color='blue',fill=NA)
```                       

Fills:
```{r}        
myMap + geom_polygon(aes(fill=val)) + geom_polygon(color='white',alpha=0.5,size=0.2)
```

`size` controls the width of the boundaries.  `alpha` sets the translucence.

### Use the with the bells and whistles of `ggplot2`.

Add in more features.  Provide a list of examples

Important one: ```coord_map(project='aitoff')``` or `'mercator'` or `'polyconic'`.  See the `mapproj` package for more info.

```{r}
myMap + geom_polygon(aes(fill=val)) + geom_polygon(color='white',alpha=0.5,size=0.2) +
  coord_map(project='polyconic')  # <---- the bell and whistle
```


* The pallete?  Or make this an argument to `makeMap()`
* Suppress axis ticks or box around plot.

### Miscellaneous

* There seem to be some points from USSR that are given in longitude west (negative) rather than longitude east (positive).  It might be a good exercise to fix them.

* More generally, the world map is out of date.  Some suggestions for fixing this are [here](http://stackoverflow.com/questions/15757793/updated-world-map-for-r-maps-package).  But there is a whole literature here (see `maptools` package) and we don't want to get too deeply involved.
> Is there a way to convert `maptools::wrld_simpl` into a format that can be used by `map_data()`?





