---
title: "Bird and Airplanes"
author: "Daniel Kaplan"
date: "April 23, 2014"
output: html_document
---

[FAA data site](http://wildlife.faa.gov/database.aspx).  You can do one airport at a time.

A [blog entry](http://genedan.com/tag/faa-strike-wildlife-strike-database/) on these data and `ggplot`
```{r include=FALSE}
# Can't read the .xls files with this package.  So, they have been
# resaved as CSV files
require(XLConnect) # Package for reading rectangular regions of XLS spreadsheets
require(dplyr)
require(lattice)
```

### Read in the FAA database

```{r}
strikesRaw <- data.frame()
for (dataFileName in c('MSP.csv','CA.csv')){
tmp <- read.csv(dataFileName)
strikesRaw <- rbind(strikesRaw,tmp)
}
```

#### Columns to keep

For the incident-by-incident data
```{r}
keepCols <- c("STATE","AIRPORT_ID","AIRPORT","OPID","ATYPE",                "SPECIES_ID","DAMAGE", "COST_REPAIRS", "AC_CLASS",              "INCIDENT_MONTH","INCIDENT_YEAR","TIME_OF_DAY",         
"TIME", "HEIGHT", "SPEED","DISTANCE","EFFECT","SKY","PRECIP", "BIRDS_SEEN", "BIRDS_STRUCK","SIZE","WARNED","INDICATED_DAMAGE")
incidents <- subset(strikesRaw,select=keepCols)
```

### FAA Tables

```{r}
damage <- data.frame(DAMAGE=c('N','M','M?','S','D','A','B','C','D'),
  DamageDesc=c('None','Minor','Uncertain','Substantial','Destroyed'
               '>$2000K','>$500K','>$50K',,'$50K.or.less' ))
```


### Extract sub-tables

```{r}
aircraft <- unique(
  subset(strikesRaw,
         select=c('ATYPE',"AC_CLASS","AC_MASS",'NUM_ENGS')
         )
  )
```

#### Airline

```{r}
airlines <- unique(subset(strikesRaw,select=c('OPERATOR','OPID')))
```

#### Airports

```{r}
airports <- unique(subset(strikesRaw,select=c('AIRPORT_ID', 'AIRPORT')))
```

#### Species

```{r}
species <- unique(subset(strikesRaw,select=c('SPECIES_ID', 'SPECIES')))
```

### Save in RDA format

```{r eval=FALSE}
save(incidents,airlines,airports,species,file='FAA_Bird_Strikes.Rda')
```

## Questions

### What are the most common types of birds?

#### Tally and arrange

**group** **tally**
```{r}
group_by(incidents,SPECIES_ID) %.% tally(sort=TRUE) -> speciesCount
```

Add in the species name: **inner join** **arrange**
```{r}
inner_join(speciesCount,species) %.% arrange(desc(n))
```

#### How are TIME and TIME_OF_DAY related?

**bwplot**

```{r}
bwplot( TIME ~ TIME_OF_DAY, data=incidents )
```

#### What altitudes?

This is a **density** problem.

```{r}
densityplot( ~HEIGHT, subset(incidents,HEIGHT>0))
```

#### Does the time of day depend on the `INCIDENT_MONTH`?

**Group** two variables, **tally**, **scatterplot** with groups

```{r}
res <- group_by(incidents,TIME_OF_DAY,INCIDENT_MONTH) %.% tally()
```

From `mPlot(res)`
```{r}
ggplot(data=res, aes(x=INCIDENT_MONTH, y=n)) + geom_point()  + aes(colour=TIME_OF_DAY)  + stat_smooth(method=loess) + theme(legend.position="top") + labs(title="")
```

### Species with Time of Day and Month

Find the n most common species:

```{r}
nspecies=10
scount <- group_by(incidents, SPECIES_ID) %.% tally() %.% 
  arrange(sort=TRUE)
topspecies <- scount[1:nspecies]
small <- select(incidents)


## ADVANCED
ADVANCED: PULL OUT THE 10 most common in each state.  Reduce this to a list of birds overall.  Plot the distribution versus MONTH, TIME, HEIGHT

