---
title: "DPLYR Style"
author: "Daniel Kaplan"
date: "April 21, 2014"
output: html_document
---


```{r include=FALSE}
require(dplyr)
require(mosaic)
```

Style for doing some basic things with `dplyr`

## Watch Out!

`plyr` also has a `summarize()` function.  

`plyr` has `join()`. For the `dplyr` versions, it must be `inner_join()`, `left_join()`, `semi_join()`, `anti_join()`.

`mosaic` and `dplyr` both have `tally()`.

## Counting

**DECIDE** whether to use `tally()`

```{r}
group_by(Galton, sex) %.% dplyr::tally( )
```

Or `summarize()` and `n()` so that names can be used and other quantities calculated at the same time?
```{r}
group_by(Galton, sex) %.% 
  dplyr::summarize( count=n(),hgt=mean(height)/median(height) )
```


## Rename Variables

### A Single variable

Use `select()` with `new=old`
```{r}
Galton <- select(Galton, child=height)
head(Galton)
```

### Change prefixes on a bunch of variables

```{r}
head(select(iris, pl = starts_with("Petal")))
```

## Joining

When joining, use only the needed columns: specify them with `select()`.  That way, you can rename them as needed for matching variables in the join.


